How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-04-15 doc/paper.tex
changes to introduction to improve clarity

The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the nea future. Such predictions are difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented. This has led to the suggestions that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}. The fossil record preserves information about the full histories, include ultimate extinction, of thousands of lineages, and this information can help to augment the shorter-term higher-resolution data used to make risk assessments of extant taxa.

Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.

This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.

Here we ask how precise risk predictions based on fossil data might be.  Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions.  However, we can address the question by taking a given point in the geological past, developing predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g. ``future'', from the point of view of the model) extinction/survival events.

Both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}. Given this variation, how accurate are our assessments based on past events likely to be when applied to the future? Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.

Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction. What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range extinction intensities and selectivities.

Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that geographic range trajectories are highly variable \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian.  In the former case, knowledge of the specific past trajectory of a species -whether its range has expanded or contracted over a given timespan- might help to improve assessments of its current risk.  In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past.  Discriminating among these alternative models is thus very important for determining how best to incorporate fossil data in present risk assessments.

@@

We also find that our four models are practically identical in their ability to make forecasts both in-sample and out-of-sample. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant. The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance.

>>>>

The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the near future. Predicting which species will go extinct is difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented. This has led to the suggestions that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}. The fossil record preserves information about the full histories, include ultimate extinction, of thousands of lineages, and this information can help to augment the shorter-term higher-resolution data used to make risk assessments of extant taxa.

Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.

Here we ask how precise risk predictions based on fossil data might be. Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions. However, we can address the question by taking a given point in the geological past, developing a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g. ``future'', from the point of view of the model) extinction/survival events.

Extinction intensity (average rate) and selectivity (difference in risk between taxa) vary through time and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}. What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? Given this variation, how accurate are our assessments based on past events likely to be when applied to the future? Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.

Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian. In the former case, knowledge of the specific past trajectory of a species -- whether its range has expanded or contracted over a given time span -- might help to improve assessments of its current risk. In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past. Discriminating among these alternative models is thus very important for determining how best to incorporate fossil data in present risk assessments.

@@

We also find that our four models are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant. The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance.
